---
title: "Neonatal mortality"
author: Ingeborg Gullikstad Hem (ingeborg.hem@ntnu.no)
output: 
  rmarkdown::html_vignette:
    toc: true
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{neonatal_mortality}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(rmarkdown.html_vignette.check_title = FALSE)
```

```{r setup}
library(makemyprior)
```


### Model

We carry out a similar study of neonatal mortality in Kenya as one by @fuglstad2020. We model the neonatal mortality,
defined as the number of deaths if infants the first month of live per birth.
We use the linear predictor:
\[
  \eta_{i,j} = \mathrm{logit}(p_{i,j}) = \mu + x_{i,j} \beta + u_i + v_i + \nu_{i,j}, \ i = 1, \dots, n, \ j = 1, \dots, m_i,
\]
and use $y_{i,j} | b_{i,j}, p_{i,j} \sim \mathrm{Binomial}(b_{i,j}, p_{i,j})$, for cluster $j$ in county $i$. We have between $m_i \in \{6, 7, 8\}
clusters in each of the $n = 47$ counties (see e.g. @fuglstad2020 for a map of the counties).

* $b_{i,j}$ is the number of live births,
* $y_{i,j}$ is the number of neonatal deaths,
* $\mu$ is an intercept with a $N(0, 1000^2)$ prior,
* $\beta$ is a coefficient with a $N(0, 1000^2)$ prior for $x_{i,j}$, which is an indicator classifying cluster $j$ in county $i$ as urban ($x_{i,j} = 1$) or rural ($x_{i,j} = 0$),
* $\nu_i \sim N_n(0, \sigma_{\nu}^2)$ is an i.i.d. random effect for cluster
* $v_i \sim N_n(0, \sigma_v^2)$ is an i.i.d. random effect for county, and
* $\mathbf{u}$ is a Besag effect on county with variance $\sigma_u^2$ and a sum-to-zero constraint.

We need a neighborhood graph for the counties, which is found in ``makemyprior``. We scale the Besag effect to have 
a generalized variance equal to $1$.

```{r}
# neighborhood graph
graph_path <- paste0(path.package("makemyprior"), "/neonatal.graph")

formula <- y ~ urban + mc(nu) + mc(v) +
  mc(u, model = "besag", graph = graph_path, scale.model = TRUE)

```

We use the dataset ``neonatal_mortality`` in ``makemyprior``, and present three priors.
We do not carry out inference, as it takes time and will slow down the compilation of the vignettes by a lot,
but include code so the user can run the inference themselves.


### Prior 1

We prefer coarser over finer unstructured effects, and unstructured over structured effects.
That means that we prefer $\mathbf{v}$ over $\mathbf{u}$ and $\mathbf{v} + \mathbf{u}$ over $\mathbf{\nu}$ in
the prior.
We achieve this with a prior that distributes the county variance with shrinkage towards 
the unstructured county effect, and the total variance towards the county effects.
Following @fuglstad, we induce shrinkage on the total variance such that we have a 
90\% credible interval of
$(0.1, 10)$ for the effect of $\exp(v_i + u_i + \nu_{i,j})$. We use 
the function ``find_pc_prior_param``
in ``makemyprior`` to find the parameters for the PC prior:


```{r}
set.seed(1)
find_pc_prior_param(lower = 0.1, upper = 10, prob = 0.9, N = 2e5)

```


```{r}
prior1 <- make_prior(
  formula, neonatal_data, family = "binomial",
  prior = list(tree = "s1 = (u, v); s2 = (s1, nu)",
               w = list(s1 = list(prior = "pc0", param = 0.25),
                        s2 = list(prior = "pc1", param = 0.75)),
               V = list(s2 = list(prior = "pc",
                                  param = c(3.35, 0.05)))))

prior1

```

```{r, fig.width = 5, fig.height = 3, eval = FALSE}
plot_prior(prior1) # or plot(prior1)
```

```{r, fig.width = 5, fig.height = 3, eval = FALSE}
plot_tree_structure(prior1)
```


Inference can be carried out by running:

```{r, eval = FALSE}
posterior1 <- inference_stan(prior1, iter = 15000, warmup = 5000,
                            seed = 1, init = "0", chains = 1)

plot_posterior_stan(posterior1, param = "prior", plot_prior = TRUE)

```

For inference with INLA:

```{r, eval = FALSE}
posterior1_inla <- inference_inla(prior1, Ntrials = neonatal_data$Ntrials)
plot_posterior_stdev(posterior1_inla)

```

Note the ``Ntrials`` argument fed to ``inference_inla``.

### Prior 2


We use a prior without any knowledge, and use the default prior:

```{r}
prior2 <- make_prior(formula, neonatal_data, family = "binomial")

prior2

```



```{r, fig.width = 5, fig.height = 3, eval = FALSE}
plot_prior(prior2)
```

```{r, fig.width = 5, fig.height = 3, eval = FALSE}
plot_tree_structure(prior2)
```


Inference can be carried out by running:

```{r, eval = FALSE}
posterior2 <- inference_stan(prior2, iter = 15000, warmup = 5000,
                            seed = 1, init = "0", chains = 1)

plot_posterior_stan(posterior2, param = "prior", plot_prior = TRUE)

```

```{r}
sessionInfo()
```

### References
